Quantum Hoare Type Theory

### 2020-10-29 ms-paper/ms-paper.tex
Simplify syntax some more

\begin{figure}[ht]
	\begin{tabular}{lrcl}
		\textit{Types} & $A, B, C$ & ::= &
		\begin{minipage}[t]{0.5\columnwidth}%
			$ \type{1} \mid \type{Qbit} \mid \type{U} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
		\end{minipage}\\ \\
		\textit{Propositions} & $P, Q, R$ & ::= &
		\begin{minipage}[t]{0.5\columnwidth}%
			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid U \cdot P \mid X \in_q S \mid X =_q \psi \mid \kw{uniform}(X) \mid \kw{separable}(X) \mid \kw{classical}(X) $
		\end{minipage}\\ \\
		% \textit{Q. Heaps} & $H, G$ & ::= & $ h \mid \kw{empty} \mid \kw{upd}(H, M, N) $\\ \\
		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K N \mid N : A $\\
		\textit{Intro terms} & $N, O$ & ::= & $ K \mid () \mid \lambda x.N \mid \kw{do} E $\\ \\
		\textit{Q. Commands} & $c$ & ::= &
		\begin{minipage}[t]{0.5\columnwidth}%
			$ \kw{init0} \mid \kw{init1} \mid \kw{apply} U \kw{to} N \mid \kw{measure} N$
		\end{minipage}\\
		\textit{Computations} & $E, F$ & ::= & $ \kw{return} N \mid x \leftarrow K; E \mid x \Leftarrow c; E \mid x =_A N; E $\\ \\
%		\textit{Variable context} & $\Delta$ & ::= & $ \cdot \mid x : A, \Delta $ \\
%		\textit{Q. Heap context} & $\Psi$ & ::= & $ \cdot \mid h, \Psi$ \\
%		\textit{Proposition context} & $\Gamma$ & ::= & $ \cdot \mid P, \Gamma$ \\
		\textit{Unitaries} & $U$ & ::= & $ I \mid X \mid Y \mid Z \mid H \mid CX \mid CZ \mid S \mid T$
	\end{tabular}
	\Description{Syntax of Quantum Hoare Type Theory}
	\caption{Syntax of Quantum Hoare Type Theory}
	\label{fig:syntax}
\end{figure}

\paragraph{Types} We include primitive types for unit, qubits, and unitary operations. There are also type constructors for pairs, dependent functions and the Hoare type from HTT.

\paragraph{Propositions}
The predicate language is inspired by \citet{unruh2019} that follows subspace semantics. We need to be explicit about what we mean when we say that a certain state satisifes a certain predicate in the quantum setting. In this model, the propositions are subspaces of the Hilbert space corresponding to a quantum state. Here are some examples:
\begin{itemize}
	\item $\top$ represents the complete Hilbert space that is always satisfied by any state.
	\item $\bot$ represents nothing, it is never satisfied.
	\item $P \wedge Q$ represent that both P and Q hold (conjunction), it is equivalent to the intersection of the subspaces.
	\item $P \vee Q$ similarly represents disjunction and is modeled as the sum (linear combination) of the subspaces P and Q.
	\item If an operator M is applied to a state satisfying a proposition P, then the new proposition can be written as $U \cdot P$.
	\item If we want to write that a variable w, or a list of variables W is in a specific state, we can model that proposition as the variable(s) being in the span of the state and use the notation, $W =_q \psi$.
\end{itemize}

Here are some examples. The postcondition of the program \texttt{qplus} below can be written as $r =_q \ket{+}$. The postcondition for the program \texttt{bell00} as $(a, b) =_q \frac{1}{\sqrt{2}}\ket{00}+\frac{1}{\sqrt{2}}\ket{11}$ or equivalently as $(a, b) =_q \ket{\Phi^+}$..

There are other propositions that involve ghosts. \todo{TODO}

>>>>

\begin{figure}[ht]
	\begin{tabular}{lrcl}
		\textit{Types} & $A, B, C$ & ::= &
		\begin{minipage}[t]{0.5\columnwidth}%
			$ \type{1} \mid \type{Qbit} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
		\end{minipage}\\ \\
		\textit{Propositions} & $P, Q, R$ & ::= &
		\begin{minipage}[t]{0.5\columnwidth}%
			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid U \cdot P \mid \textbf{X} =_q \psi \mid \kw{uniform}(\textbf{X}) \mid \kw{separable}(\textbf{X}) \mid \kw{classical}(\textbf{X}) $
		\end{minipage}\\ \\
		% \textit{Q. Heaps} & $H, G$ & ::= & $ h \mid \kw{empty} \mid \kw{upd}(H, M, N) $\\ \\
		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K N \mid N : A $\\
		\textit{Intro terms} & $N, O$ & ::= & $ K \mid () \mid \lambda x.N \mid \kw{do} E $\\ \\
		\textit{Q. Commands} & $c$ & ::= &
		\begin{minipage}[t]{0.5\columnwidth}%
			$ \kw{init0} \mid \kw{init1} \mid \kw{apply} U \kw{to} N \mid \kw{measure} N$
		\end{minipage}\\
		\textit{Computations} & $E, F$ & ::= & $ \kw{return} N \mid x \leftarrow K; E \mid x \Leftarrow c; E \mid x =_A N; E $\\ \\
%		\textit{Variable context} & $\Delta$ & ::= & $ \cdot \mid x : A, \Delta $ \\
%		\textit{Q. Heap context} & $\Psi$ & ::= & $ \cdot \mid h, \Psi$ \\
%		\textit{Proposition context} & $\Gamma$ & ::= & $ \cdot \mid P, \Gamma$ \\
		\textit{Unitaries} & $U$ & ::= & $ I \mid X \mid Y \mid Z \mid H \mid CX \mid CZ \mid S \mid T$
	\end{tabular}
	\Description{Syntax of Quantum Hoare Type Theory}
	\caption{Syntax of Quantum Hoare Type Theory}
	\label{fig:syntax}
\end{figure}

\paragraph{Types} There are primitive types for unit and qubits. There are also type constructors for pairs, dependent functions and the Hoare type from HTT.

\paragraph{Propositions}
The predicate language is inspired by \citet{unruh2019} that follows subspace semantics. We need to be explicit about what we mean when we say that a certain state satisifes a certain predicate in the quantum setting. In this model, the propositions are subspaces of the Hilbert space corresponding to a quantum state. We use the metavariable \textbf{X} to refer to a list of quantum variables. Here are some examples:
\begin{itemize}
	\item $\top$ represents the complete Hilbert space that is always satisfied by any state.
	\item $\bot$ represents nothing, it is never satisfied.
	\item $P \wedge Q$ represent that both P and Q hold (conjunction), it is equivalent to the intersection of the subspaces.
	\item $P \vee Q$ similarly represents disjunction and is modeled as the sum (linear combination) of the subspaces P and Q.
	\item If a unitary operator U is applied to a state satisfying a proposition P, then the new proposition can be written as $U \cdot P$.
	\item If we want to write that a list of variables \textbf{X} is in a specific state, we can model that proposition as the variables being in the span of the state and use the notation, $\textbf{X} =_q \psi$.
\end{itemize}

Here are some examples. The postcondition of the program \texttt{qplus} below can be written as $r =_q \ket{+}$. The postcondition for the program \texttt{bell00} as $(a, b) =_q \frac{1}{\sqrt{2}}\ket{00}+\frac{1}{\sqrt{2}}\ket{11}$ or equivalently as $(a, b) =_q \ket{\Phi^+}$.

There are other propositions that involve ghosts:
\begin{itemize}
	\item uniform(\textbf{X}) is an abbreviation for $(\textbf{X}, \textbf{e}) = \frac{1}{\sqrt{2}}(\ket{00}+\ket{11})$ where \textbf{e} is  a fresh entangled ghost variable. It means that X is uniformly distributed.
	\item separable(\textbf{X}) is an abbreviation for $\textbf{X} \equiv_{q} \textbf{u}$ where \textbf{u} is a fresh unentangled ghost variable.
	\item class(\textbf{X}) is an abbreviation for $\textbf{X} \equiv_{cl} \textbf{u}$ where \textbf{u} is a fresh unentangled ghost variable. The classical equality symbol means that the variables on both side will measure to be the exact same values. That is there is no superposition involved.
\end{itemize}
